Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap

Now that we have a semantics for \isotopessa programs, we can show that it
matches the syntax by showing that the syntactic and semantic notions of
substitution coincide. We begin by giving a semantics for the judgement
\(\issubst{\gamma}{\Theta}{\Gamma}\), which we will interpret as a pure morphism
taking \(\dnt{\Theta}\) to \(\dnt{\Gamma}\), defined via the rules in Figure
\ref{fig:ssa-subst-semantics}, which can simply be viewed as computing all
\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\). In particular, we interpret
\rle{subst-nil} as the identity function, and \rle{subst-cons} by splitting the
context via \(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), and then
passing \(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) and \(\dnt{\Theta_x}\) through
\(\dnt{\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}\) to yield \(\dnt{A}\). We give a
similar semantics for label-substitutions: \rle{lb-subst-nil} is interpreted as
the zero morphism \(0_{\dnt{\ms{K}}}: \mc{C}_0^\varnothing(\mb{0},
\dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in the coproduct \(\dnt{\ms{L}}
+ \dnt{\Gamma} \otimes \dnt{A}\), sending the former to
\(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the denotation of
\(\mc{L}(\lbl{\ell}, x)\).

@@

Given this, we can now state the theorem of semantic substitution, which
essentially states that the semantics of a substituted term or block is the same
as the semantics of that term or block preceded by let-bindings corresponding to
the variables in the substitution, which is itself the same as the semantics of
the original term or block precomposed with the semantics of the substitution as
given in Figure \ref{fig:ssa-subst-semantics}. Similarly, the semantics of a
\textit{label-substituted} block is the same as the semantics of the original
block \textit{postcomposed} with the semantics of that label-substitution.
Stated formally, this becomes:
\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}}
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \]
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}  
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  \begin{itemize} 
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
    \[
      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
    \]
    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
    \[
      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
    \]
    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  \end{itemize}
\end{theorem}

>>>>

Now that we have a semantics for \isotopessa programs, we can show that it
matches the syntax by showing that the syntactic and semantic notions of
substitution coincide. We begin by giving a semantics for
\(\issubst{\gamma}{\Theta}{\Gamma}\) in Figure \ref{fig:ssa-subst-semantics},
which we will interpret as a pure morphism taking \(\dnt{\Theta}\) to
\(\dnt{\Gamma}\). This morphism can simply be viewed as computing all
\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\); in particular, we interpret
\rle{subst-cons} by splitting the context via
\(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), passing
\(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) to compute the rest of the
\(\gamma(y)\) in \(\Gamma\) recursively, and using \(\dnt{\Theta_x}\) to compute
\(\gamma(x)\). Similarly, we interpret label-substitutions as morphisms from \(\dnt{\ms{L}}\) to
\(\dnt{\ms{K}}\) which, for each input label in \(\ms{L}\), performs a
computation targeting \(\ms{K}\). In particular, \rle{lb-subst-nil} is
interpreted as the zero morphism \(0_{\dnt{\ms{K}}}:
\mc{C}_0^\varnothing(\mb{0}, \dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in
the coproduct \(\dnt{\ms{L}} + \dnt{\Gamma} \otimes \dnt{A}\), sending the
former to \(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the
denotation of \(\mc{L}(\lbl{\ell}, x)\).

@@

We can now state the theorem of semantic substitution, which essentially states
that the semantics of a substituted term or block is the same as the semantics
of that term or block preceded by let-bindings corresponding to the variables in
the substitution, which is the same as the semantics of the original term or
block precomposed with the semantics of the substitution. Similarly, the
semantics of a \textit{label-substituted} block is the same as the semantics of
the original block \textit{postcomposed} with the semantics of that
label-substitution. Stated formally, we obtain:
\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \(
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}}
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \)
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{\exprletsubst{\substctx{\gamma}{\Gamma}}}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}  
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  \begin{itemize} 
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
    \(
      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
    \)
    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
    \[
      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
    \]
    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  \end{itemize}
\end{theorem}
